Chaotic mixing of tidal debris

### 2015-06-21 ms.tex
more words

\todo{apw}{describe other features in lyapunov grid and unstable resonances, etc.}
The strongest chaotic regions appear in both of the above grids (Figure~\ref{fig:lyapmap} and Figure~\ref{fig:freqdiff}). The resonant structure is revealed in the frequency diffusion map from integrations of just 80 orbital periods, though the majority of the orbits have estimated chaotic timescales corresponding to thousands of orbital periods. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids and compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen at random from a single column of the orbit grid based on their frequency diffusion time, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on these three orbits. The initial mass of the clusters is \todo{kvj}{2.5E4 \msun contained within what again? 35 scale radii? but this is plummer??}. For each simulation, the progenitor is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and evolve for $\approx$16 orbital periods so that each simulation finishes again at subsequent apocenter. Figure~\ref{fig:nbodysims} shows the final particle distributions rotated so that the angular momentum of the progenitor orbit is aligned with the $z$-axis. A thin stream forms on the regular orbit (left column), a more diffuse stream on the mildly chaotic orbit (middle column), and a ``fanned'' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that stream formed on this orbit appears to be more diffuse than the regular orbit stream.

@@

We have considered here a simple triaxial gravitational potential chosen to mimic the median properties of dark matter halos formed in dark-matter-only simulations (or the large-scale properties of halos formed in simulations with baryonic effects). We have found that ``mean'' chaos indicators such as the Lyapunov exponent or frequency diffusion rate predict that chaos should be dynamically unimportant in the halos of galaxies for the majority of orbits (e.g., the predicted timescales are 100-1000s of times longer than the typical orbital times, at least in this simple potential). However, while the mean chaotic evolution of, for example, the fundamental frequencies of an orbit may be prohibitively long, the frequencies of even mildly chaotic orbits do evolve on a smaller scale over much shorter times. The magnitude of these variations depends on the resonant structure local to the orbit (which is highly potential-dependent), but this result is generic to any potential with resonances and regions of stochasticity. When this small-scale but short-time frequency evolution has an amplitude comparable to or larger than the intrinsic spread in orbital frequencies for tidal debris, the debris will appear dynamically hotter and diffuse faster than it would along a regular orbit. We have found that even in the simple potential used in this work (Section~\ref{sec:potential}), a large fraction of orbits would naively be considered weakly chaotic, but tidal streams formed on these orbits would appear morphologically different (e.g., more diffuse) than those formed on regular orbits.

If this is the case for a single-component, static, smooth potential, the mere existence of thin tidal streams in the halo of the Milky Way either (1) provides useful information about the potential on these scales by, e.g., implying a large degree of regularity, or (2) indicates that the thin, long streams (GD-1, ??) are on regular orbits. These are not mutually exclusive --- in fact, if the streams are on regular orbits, this would be a powerful way to check or rule out possible potential models. The longest thin streams are most valuable for this effort because they have clearly evolved for a long time, but the debris remains compact in frequency space. For shorter thin streams it will be hard to decouple the unknown evolution time from the analysis.

@@

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx $, $z \approx $, the orbit of this stream likely passes through the MW disk, feels the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and has a short orbital period (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if stars were disrupted on previous pericentric passages, this debris is likely ``fanned'' but still near the observed portion of the stream. The fanned debris will be much lower surface brightness and thus much more difficult to detect, but would be [...].

\subsection{Potential recovery with streams}

Recovering the potential of triaxial systems is not trivial. Danger of orbit fitting because chaotic orbits can reproduce morphologies of regular orbits.

\subsection{Potential choice}
[discussion of potential choice?]
Why not disk + bulge? What do we expect when we add?

@@

The potential considered in this work is ``unrealistic'' for the potential of a Milky-Way-like galaxy in that it is static, smooth, and does not contain baryonic components. Hydrodynamic simulations of forming galactic disks in cosmological dark matter halos have shown that baryonic feedback and relaxation can significantly change the inner distribution of dark matter and either make the potential more spherical or oblate \citep{many}.

@@

We have shown that ...

@@

% Figure ??
\clearpage
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/ensemble-map.png}
\caption{The same grid of initial conditions as Figure~\ref{fig:lyapmap} and other previous figures, but now we consider the density evolution of small ensembles of orbits around each of the orbits in this grid. We generate globular-cluster-scale ensembles of orbits around each ``parent'' orbit in the grid (Section~\ref{sec:results2}) and follow the density of the ensemble as a function of time using kernel density estimation (KDE) with an adaptive Epanechnikov kernel. Three orbits are highlighted with square markers --- from top to bottom, these are the strongly chaotic (S), weakly chaotic (W), and regular (R) orbits of Table~\ref{tbl:three-orbits} (see Section~\ref{sec:results2}).} 
\label{fig:ensemblemap-meandensity}
\end{center}
\end{figure*}

@@

% Figure ??
\clearpage
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/freq-evolution.png}
\caption{ } 
\label{fig:three-orbits-freqs}
\end{center}
\end{figure*}

>>>>

The strongest chaotic regions (black) appear in both of the above grids (Figure~\ref{fig:lyapmap} and Figure~\ref{fig:freqdiff}). Some of the weakly chaotic unstable resonances do appear in the Lyapunov time map, especially where resonances overlap --- for example, near $(x,z) \approx(13,7)$ and $(x,z) \approx(31,13)$ where there is a slight hint of weak chaos (grey) in Figure~\ref{fig:lyapmap}. The details of the resonant structure is revealed in the frequency diffusion map from integrations of just 80 orbital periods, though the majority of the orbits have estimated chaotic timescales corresponding to thousands of orbital periods. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids and compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen at random from a single column of the orbit grid based on their frequency diffusion time, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on these three orbits. The initial mass of the clusters is \todo{kvj}{2.5E4 \msun contained within what again? 35 scale radii? but this is a plummer sphere?}. For each simulation, the progenitor is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and evolve for $\approx$16 orbital periods so that each simulation finishes again at subsequent apocenter. Figure~\ref{fig:nbodysims} shows the final particle distributions rotated so that the angular momentum of the progenitor orbit is aligned with the $z$-axis. A thin stream forms on the regular orbit (left column), a more diffuse stream on the mildly chaotic orbit (middle column), and a ``fanned'' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that stream formed on this orbit appears to be more diffuse than the regular orbit stream.

@@

We have considered here a simple triaxial gravitational potential chosen to mimic the median properties of dark matter halos formed in dark-matter-only simulations (or the large-scale properties of halos formed in simulations with baryonic effects). We have found that ``mean'' chaos indicators such as the Lyapunov exponent or frequency diffusion rate predict that chaos should be dynamically unimportant in the halos of galaxies for the majority of orbits (e.g., the predicted timescales are 100-1000s of times longer than the typical orbital times, at least in this simple potential). However, while the mean chaotic evolution of, for example, the fundamental frequencies of an orbit may be prohibitively long, the frequencies of even mildly chaotic orbits do evolve on a smaller scale over much shorter times. The magnitude of these variations depends on the resonant structure local to the orbit (which is highly potential-dependent), but this result is generic to any potential with resonances and regions of stochasticity. When this small-scale but short-time frequency evolution has an amplitude comparable to or larger than the intrinsic spread in orbital frequencies for tidal debris, the debris will appear dynamically hotter and become diffuse faster than along a regular orbit. We have found that even in the simple potential used in this work (Section~\ref{sec:potential}), a large fraction of orbits would naively be considered weakly chaotic, but tidal streams formed on these orbits would appear morphologically different (e.g., more diffuse) than those formed on regular orbits.

If this is the case for a single-component, static, smooth potential, the mere existence of thin tidal streams in the halo of the Milky Way either (1) provides useful information about the potential on these scales by, e.g., implying a large degree of regularity, or (2) indicates that the thin, long streams (e.g., GD-1) are on regular orbits. These are not mutually exclusive --- in fact, if the streams are on regular orbits, this would be a powerful way to check or rule out possible potential models by requiring that the progenitor orbits remain regular. The longest thin streams are most valuable for this effort because they have clearly evolved for a long time, but the debris remains compact. For shorter thin streams it will be hard to decouple the unknown evolution time from enhanced density evolution from chaos.

@@

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx 1$ kpc, $z \approx 5$ kpc, the orbits of the stream stars likely pass through the MW disk, feel the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and have short orbital periods (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if a significant number of stars were disrupted on previous pericentric passages, this older debris may be ``fanned'' and still near the observed portion of the stream. The fanned debris would be much lower surface brightness and thus much more difficult to detect. However, the detection of such stars would confirm that enhanced density evolution is a dynamically relevant process for thin streams in real galaxies.

\subsection{The dangers of stream modeling}

Tidal stream modeling is one of the most promising ways to constrain the 3D mass distribution around the Milky Way at distances of 10s to 100s of kpc. Methods that use tidal streams to infer properties of the Galactic potential typically operate by constructing models of the debris distribution using either the present-day phase-space density \citep[e.g.,][]{kuepper12, kuepper15}, time-of-disruption phase-space density \citep{apw14}, or the density in angle-action coordinates \citep[e.g.,][]{sanders14, bovy14}. All of these methods may fail for modeling globular-cluster streams on mildly chaotic orbits.

Consider a globular cluster system that disrupts along a mildly chaotic orbit so that a significant portion of the resulting stream is more diffuse (lower surface brightness) and therefore unobservable due to observational biases. The high-density part of the stream may still appear thin (e.g., the recently disrupted debris in the middle panels of Figure~\ref{fig:nbodysims}). If the stream modeling focuses on reproducing the present-day density of the stream and the potential is restricted to an integrable, analytic form, the debris may be well fit by a particular orbit and evolution time in such a potential. However, the inferences about the potential or mass distribution may be biased significantly and the inferences about the evolution time and internal properties of the system will certainly be incorrect.

[what happens in angle-action coordinates?]

[what about opposite effect -- if potential model too flexible, can always find chaotic orbit to reproduce regular orbit?]

\subsection{Potential choice}

The potential considered in this work is ``unrealistic'' for the total potential of a Milky-Way-like galaxy in that it is static, smooth, and does not contain baryonic components. Hydrodynamic simulations of forming galactic disks in cosmological dark matter halos have shown that baryonic feedback and relaxation can significantly change the inner distribution of dark matter and either make the potential more spherical or oblate \citep{many}. However, the significance of baryonic relaxation or of time-dependence, triaxiality, and substructure on shaping the matter distribution within the Milky Way is largely unknown. We now consider each of these effects independently and comment on the expected consequences of these perturbations.

\subsubsection{Baryonic components}

\cite[][D08]{debattista08} and \cite[][V10]{valluri10} studied the orbit evolution induced by growing a baryonic disk in dark matter halos with various shapes and orientations (e.g., prolate, triaxial). In the simulations of D08, the disk component is artificially fixed along one of the principal axes of the halo potential and the mass of the disk is grown adiabatically. D08 found that (in general) the growth of a disk slowly deforms the orbits of mass tracers (e.g., dark matter particles) and preferentially populates tube orbits but does not result in a large degree of chaotic scattering. Consequently, the inner shape of the potential becomes more oblate or spherical due to the prevalence of tube orbits.

If the inner halos of galaxies are indeed close to spherical or oblate, the majority of orbits will be regular and chaos will be less important, however this is far from conclusive. First, there are observational hints that the intrinsic alignment of galactic disks may be close to random \citep{??} (rather than fixed to symmetry planes of the host halo potential). Simulations since the work of D08 have shown that by ``freeing'' the disk component, the angular momentum vector of the disk can precess and wander erratically away or around the principal axes of the host halo \citep{debattista15}. It is unknown what effect this would have on the orbit structure of nearby halo orbits, but we naively expect that this would destabilize a significant number of orbits. Second, if the outer potential remains triaxial, the transition region between, e.g., oblate and triaxial may be populated by chaotic orbits. We have found from simple experiments in superposing potential components that transition regions between potential shapes induce significant amounts of chaos. Though superposing potentials is not realistic, it is at least suggestive that such configurations might be unstable. A more careful exploration of this phenomenon will be presented in future work. Lastly, the existence of a bulge or thin disk component may enhance chaos through a process similar to periodic forcing of the orbit. Passage through a thin disk results in a nearly instantaneous, impulsive forcing of the orbit at a frequency set by the circulation frequency of the orbit. This may act to increase the significance of chaos along disk-crossing orbits \citep[e.g.,][]{kandrup00}. This effect is likely unresolved in many simulations where the force and spatial resolution is insufficient to capture this effect.

\subsubsection{Subhalos}

noise in hamiltonian systems: \citep{kandrup00}

\subsubsection{Time-dependence}

[discussion of potential choice?]
Why not disk + bulge? What do we expect when we add?

@@

@@

We have shown that chaos may have a significant effect on the density evolution of tidal debris.

@@

% Figure ??
\clearpage
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth, trim={0 1cm 0 0}]{figures/ensemble-map.png}
\caption{The same grid of initial conditions as Figure~\ref{fig:lyapmap} and other previous figures, but now we consider the density evolution of small ensembles of orbits around each of the orbits in this grid. We generate globular-cluster-scale ensembles of orbits around each ``parent'' orbit in the grid (Section~\ref{sec:results2}) and follow the density of the ensemble as a function of time using kernel density estimation (KDE) with an adaptive Epanechnikov kernel. Three orbits are highlighted with square markers --- from top to bottom, these are the strongly chaotic (S), weakly chaotic (W), and regular (R) orbits of Table~\ref{tbl:three-orbits} (see Section~\ref{sec:results2}).} 
\label{fig:ensemblemap-meandensity}
\end{center}
\end{figure*}

@@

% Figure ??
\clearpage
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/freq-evolution.png}
\caption{Evolution of the fundamental frequency ratios, $r_{1,3} = \Omega_1 / \Omega_3$ and $r_{2,3} = \Omega_2 / \Omega_3$, computed over 256 orbital periods for the three orbits, R, W, S. Fundamental frequencies are computed with a window width of $\approx$40 orbital periods, and the window is shifted by one orbital period between each computation (each grey point represents the fundamental frequency ratios computed in a single window). Plotted are the per cent deviations of the frequency ratios from the value in the initial window --- the initial value is shown as a green plus sign and the final value is shown as a red x. For the regular orbit (R), the fractional variation is around $10^{-6}$ and thus all points overlap on this scale. The weakly chaotic orbit (W) displays frequency variations comparable to the spread in frequencies in globular-cluster-like tidal debris. The frequency variations of the strongly chaotic orbit (S) can be many 10s of per cent.} 
\label{fig:three-orbits-freqs}
\end{center}
\end{figure*}
